---
title: "Prediciotn of Activity Patterns Using a Random Forest Model "
author: "Dong Liang"
date: "May 16, 2016"
output: 
  html_document: 
    theme: cerulean
---




```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

```

```{r}
## Load up packages -------------------------------------------------------------------
library(caret)
library(ggplot2)
library(magrittr)
library(data.table)
library(dplyr)
library(doMC)
library(DescTools)
```





```{r data_retrieval, include = F}
mod = readRDS("/Users/DL/Documents/R project/Github-DongL/Cousera—Practical Machine Learning/model_rf.RDS")
# saveRDS(mod, 'model_rf.RDS')
```




## Background
The report is generated to identify patterns of activities when people perfom excercises. The data used in this analsyis were 

collected from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. 

In the experiment, the 6 participants were asked to perform unilateral dumbbell biceps curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E). The human activity recognition device recoded data from accelerometers on the belt, forearm, arm, and dumbell. 

Read more: http://groupware.les.inf.puc-rio.br/har#ixzz49bb9umaP

when wearing XXX which recored



## Data Acquisition
In order to predict the pattern of acitvities when the particpants did the exercise, a random forest model has been constructed using training dataset. The train and test datasets are downloade directly from the weblink as follows: [Training](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv), [Testing]("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv").

```{r}
## Load up train and test datasets -------------------------------------------------------------------
train = read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
test = read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")
```

## Data Cleaning and Preprocessing

```{r}
# Find the variable columes that contain more that 80% NAs or empty values. Those data may probably not be useful for modeling due to the limited info provided. 
setDT(train)
setDT(test)
na_gt_80perc = train[, lapply(.SD, function(c) ((is.na(c) | c == "" ) %>% sum)/length(c)) > 0.8]

# Remove the index variable - X, outcome varaible(Y) - classe and the problem id - problem_id from train and test datasets 
tr_x = train[, c(!na_gt_80perc), with = F] %>% select(-classe, -X) 
te_x = test[, c(!na_gt_80perc), with = F] %>% select(-problem_id, -X) 
tr_Y = train[, c(!na_gt_80perc), with = F]$classe


```

The original train dataset has `r dim(train)[2]` variables, but unfortunately, not all of the features recorded are useful for building a random forest model due to the high percentatge of NA and empty values. Besides, the index varaible- X and outcome variable - classe should also be excluded from predictor features. After clearning up, the remaining features to be used for model construction are: *`r names(tr_x)`*  


## Train a random forest model
I first use the trimmed train dataset and further split it into sub_train and sub_test using the `createDataPartition` function in the caret package. The sub_train dataset is used to train a random forest model with 5 fold cross validation.

```{r}
# Data preperation
intrain = createDataPartition(train$X, p = .75) %>% unlist
sub_train_x = tr_x[intrain,]
sub_train_Y = tr_Y[intrain]
sub_test_x = tr_x[-intrain, ]
sub_test_Y = tr_Y[-intrain]

# Register multi-cores using DoMC package 
registerDoMC(cores = detectCores() - 1)  

# Train a random forest model  
trCtl = trainControl(method = "cv", number = 5, allowParallel = T) 
# mod = train(sub_train_x, sub_train_Y, method = 'rf', trControl = trCtl )

```

## Model interpretation
The variable importance is shown as follows: 
```{r}
varImp(mod)
varImpPlot(mod$finalModel, type = 2)

mod$finalModel$importance
```


## Prediction on the sub_test dataset
```{r}
# Prediction
sub_test_pred = predict(mod, sub_test_x)

# Confusion Matrix
confusionMatrix(sub_test_pred, sub_test_Y)
```

I further predict the outcomes of sub_test dataset and estmate the out of sample error using `confusionMatrix` in caret package. As shown above, the trained random forest model successfully predicts the classe varaible with an overall accuracy of 0.999. 


## Prediction on the test dataset
```{r}
# Complement the lacking level information to the 20 test cases
setDF(te_x); setDF(tr_x)
for(i in seq_along(tr_x)) {levels(te_x[, i]) = levels(tr_x[, i])}

# Prediction on the test dataset      
test.pred = predict(mod, newdata = te_x)
```









